Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-05-23 paper/main.tex
more revisions

Notice, in the center/right panels of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c$ for 
the increase in precision from the low- to high- resolution output is also 
estimated with this regression model. We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the $n\times n$ diagonal matrices with elements
\[
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\]
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.
In $\Lambda_\ell$, the multiplication by 16 accounts for the sample size of the
low resolution runs.
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectra, the weighted average fluctuates too drastically due 
to the limited number of model runs.  It fails to account for the
spatial dependence inherent in the smooth but stochastic spectra.
To account for these modeling challenges, we need a more complex model which can
also provide UQ.

We consider a training-testing split with 111 cosmologies used for estimation and 6 
cosmologies held out for testing.  Our ultimate objective is to effectively predict
the underlying power matter spectra for the held-out cosmologies based on their 
8-dimensional input configuration (Section \ref{sec:pred}).  But first, we must 
obtain effective estimates of the underlying spectrum for each training 
cosmology (Section \ref{sec:hm_fit}). \textcolor{red}{(@Steve - maybe this is a
good place to introduce the ``hold out'' and ``within sample'' terms?)}

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology,
independent of the others. In the previous seciton, we used lowercase $\bar{y}$ to denote 
the observed weighted average; here, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model. Specifically, our input $X$ is the 
$n$ values of $\log_{10}(k)$ and the output is the weighted average $\bar{Y}$. 
Let $D = \{X, \bar{Y}, \Lambda\}$ represent the aggregated 
data for a particular cosmology.  Let $S$ represent the true (but unknown) underlying
power matter spectra from which $D$ was generated.
\textcolor{red}{(@Annie You suggested removing $\Lambda$ from $D$; is fixed and 
doesn't change here, but it would if $X$ or $Y$ changed. Also, we use $\Lambda$ to pre-scale the data)}

\subsection{Gaussian Process Model}

Here, we introduce the Bayesian Gaussian Process (GP) model. We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$. From the data, we obtain $\bar Y$, which is a 
noisy version of $S$.

@@

\subsection{Deep Gaussian Process Model}

>>>>

Notice, in the center/right panels of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c$ for 
the increase in precision from the low- to high- resolution output is also 
estimated with this regression model. We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the $n\times n$ diagonal matrices with elements
\[
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\]
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.
In $\Lambda_\ell$, the multiplication by 16 accounts for the sample size of the
low resolution runs.
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectrum, the weighted average fluctuates too drastically due 
to the limited amount of data.  It fails to account for the
spatial dependence inherent in the smooth but stochastic spectra,
which is pivotal to effectively inferring the true spectrum.

%We consider a training-testing split with 111 cosmologies used for estimation and 6 
%cosmologies held out for testing.  Our ultimate objective is to effectively predict
%the underlying power matter spectra for the held-out cosmologies based on their 
%8-dimensional input configuration (Section \ref{sec:pred}).  But first, we must 
%obtain effective estimates of the underlying spectrum for each training 
%cosmology (Section \ref{sec:hm_fit}). \textcolor{red}{(@Steve - maybe this is a
%good place to introduce the ``hold out'' and ``within sample'' terms?)}

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the $n$-vector of $\log_{10}(k)$ values.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  $\bar{Y}$ is typically pre-processed
in a domain-specific manner, such as that defined in Section \ref{sec:data}.
Together, $X$ and $\bar{Y}$ comprise the training data for a particular cosmology.
We will reserve specific integration 
of expert knowledge for the CAMB and Mira-Titan suites, such as this pre-processing,
for Sections \ref{subsec:camb} and \ref{subsec:mira_fit}, respectively.

\subsection{Deep Gaussian Process Hierarchical Model}

Here, we introduce the Bayesian Gaussian Process (GP) model. We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$. From the data, we obtain $\bar Y$, which is a 
noisy version of $S$.  Let $S$ represent the true (but unknown) underlying
power matter spectra from which $D$ was generated.

@@
